---
title: "Epi Machine Learning Final Project"
author: "Sydney Ng UNI: sn2863"
date: "Due: Monday, April 19, 2021 at 5:00 PM"
output: 
  html_document: 
    toc: TRUE
    toc_float: TRUE
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Dietary patterns using unsupervised analysis

### Question 1: Research Question
What are the patterns in dietary consumption during pregnancy in women and 3-4 years postnatal in children in the birth cohort study?

This is a **description** research question. A hierarchical clustering unsupervised algorithm is most appropriate here because I am most interested in examining the patterns in the dietary consumption features and understanding these clusters in terms of the inputs. This algorithm is relatively simple to visualize and interpret.

### Question 2: PCA and Hierarchical Clustering

```{r load_packages_data, message=FALSE}
library(tidyverse)
library(caret)
library(rpart.plot)
library(Amelia)
library(cluster)
library(factoextra) #scree
library(ggbiplot) #ggbiplot

rm(list = ls())
load("./exposome.RData") # Load data using path of where file is stored
```


```{r warning=FALSE}
diet_df <- read_csv("./diet_data.csv") %>% select(-ID)

# There are no missing values in the data

# Checking correlations
correlations <- cor(diet_df, use="complete.obs")
high_correlations <- findCorrelation(correlations, cutoff=0.4) 

# Removing 3 variables: 
# legume consumption during pregnancy
# total cereal consumption postnatal
# processed meat consumption postnatal
dief_df <- diet_df %>% select(-c(h_legume_preg_Ter,hs_total_cereal_Ter,hs_proc_meat_Ter))


# Centering and scaling is a good idea because the means and SDs do not seem similar
colMeans(diet_df, na.rm=TRUE)
apply(diet_df, 2, sd, na.rm=TRUE)
set.up.preprocess <- preProcess(diet_df, method = c("center", "scale"))

# Output pre-processed values
transformed_df <- predict(set.up.preprocess, diet_df) %>%
  select(-c(h_legume_preg_Ter,hs_total_cereal_Ter,hs_proc_meat_Ter))

# ----------------------------------------------------
# Implementing a PCA for feature
diet_pca <- prcomp( ~., data=transformed_df, center=TRUE, scale=TRUE,) 
  # only the features we are interested in reducing, not the outcome

#Can compare sds used to scale with the sds above to ensure they are close.
diet_pca$scale

#Generates scree plot
fviz_eig(diet_pca)

#view results of pca. Note the first three components are needed to explain at least 75% of the variance
summary(diet_pca)
  # we are more concerned with the cumulative proportion...
  # or can look at standard deviation -- cut off at 1, which is at PC9

#Identify how features loaded on the different components
diet_pca_df <- diet_pca$rotation[,1:16]
ggbiplot(diet_pca, choices=c(16,17))

# ----------------------------------------------------
# Implementing the hierarchical clustering
set.seed(100) # setting for replicability

# Create function to use within clusGap to identify the optimal number of clusters
hclusCut <- 
  function(x, k) list(cluster = cutree(hclust(dist(x, method="euclidian"), method="complete"), k=k))
gap_stat <- clusGap(transformed_df, # replace with PCA data frame here?
                    FUN = hclusCut, K.max = 10, B = 50)

# Identifying 5 as the optimal number of clusters
fviz_gap_stat(gap_stat)

clusters.hcut <- hcut(transformed_df, k=3, hc_func="hclust", hc_method="complete", hc_metric="euclidian")
clusters.hcut$size

fviz_cluster(clusters.hcut)
```


### Question 3: Outputs

Describe the outputs of the analysis in terms of their composition of the input features.

```{r describing_clustering}
input.feature.vals <- 
  bind_cols(diet_df, cluster=clusters.hcut$cluster)

input.feature.vals %>%
  group_by(cluster) %>%
  summarise_all(mean)
```


# Part 2: Choose your own supervised adventure

### Research Question:
What are the lifestyle and built environment features that best predict childhood obesity in children 6-11 years old?

My final project rationale is hypothesis generating about how lifestyle and built environment factors contribute to childhood obesity to inform future research on how individuals' day-to-day surrounding built environments can interact with their lifestyles. I will be using a Random Forest for classification.

### Data Cleaning and Processing

```{r cleaning_processing, message=FALSE, warning=FALSE}
studydata <- merge(exposome,phenotype,by="ID") %>% 
  merge(covariates, by="ID") %>% 
  mutate(hs_asthma = as.factor(hs_asthma)) %>%
  select(-ID)

builtenv_lifestyle_vars <- 
  codebook %>% filter(family %in% c("Lifestyle", "Built environment")) %>% rownames()

builtenv_lifestyle <- 
  studydata %>% 
  select(builtenv_lifestyle_vars) %>%
  mutate(pet_allergen = case_when(hs_pet_cat_r2_None == "0" ~ 0,
                                  hs_pet_cat_r2_None == "1" ~ 1,
                                  hs_pet_dog_r2_None == "0" ~ 0,
                                  hs_pet_dog_r2_None == "1" ~ 1,
                                  hs_pet_None == "No" ~ 0,
                                  hs_pet_None == "Yes" ~ 1)) %>%
  select(-c(hs_pet_cat_r2_None, hs_pet_dog_r2_None, hs_pet_None))

# Correlations
dat_num <- builtenv_lifestyle %>% select(where(is.numeric))
dat_not_num <- builtenv_lifestyle %>% select(!where(is.numeric))

correlations <- cor(dat_num, use="complete.obs")
high_correlations <- findCorrelation(correlations, cutoff=0.5) 

# Removing 16 highly correlated columns based off a correlation cutoff of 0.5
dat_num <- dat_num[-high_correlations] 

# Centering and Scaling
set.up.preprocess <- preProcess(dat_num, method = c("center", "scale"))

# Output pre-processed values
transformed_vals <- predict(set.up.preprocess, dat_num)

# Checking overlapping categorical variables
summary(dat_not_num)
transform_df <- bind_cols(transformed_vals, dat_not_num) # 45 variables

```

### Implementing a Random Forest
```{r random_forest}

```




